---
title: "Day 2 foreach demonstration"
author: "High Performance R 2025"
format: html
editor: visual
---

## Getting started with parallel R: `foreach`

Checking how many cores there are available in the session:

```{r}
parallelly::availableCores()
```

**Important:** do not use parallel::detectCores() on Roihu - it always returns the maximum number of cores in a node (384), no matter how many have been reserved. This causes problems!

```{r}
# do not use in scripts or R jobs on Roihu - this gives the wrong number of cores!
parallel::detectCores()
```

Let's start from a simple for loop and check how long it runs:

```{r}
library(tictoc)

tic()
for (i in 1:3) {
  sqrt(i)
  Sys.sleep(5) # added to make this example script run longer
}
toc()
```

Same example as above but this time using the package `foreach` (still sequential, not parallel):

```{r}
library(foreach)
library(tictoc)

tic()
foreach(i = 1:3, .combine = 'c') %do% {
  sqrt(i)
  Sys.sleep(5)
}
toc()
```

Running the same thing **in parallel** taking advantage of **3 cores**:

```{r}

library(doParallel)
registerDoParallel(cores = 3) #registering a backend for foreach

tic()
foreach(i = 1:3, .combine = 'c') %dopar% {
  sqrt(i)
  Sys.sleep(5)
}
toc()

# unregistering the backend by changing back to sequential:
registerDoSEQ()
```

What would happen without the sleep step (= a very short run)? Why?

```{r}
# sequential
tic()
foreach(i = 1:3, .combine = 'c') %do% {
  sqrt(i)
}
toc()
```

```{r}
# parallel
tic()
doParallel::registerDoParallel(cores = 3)
foreach(i = 1:3, .combine = 'c') %dopar% {
  sqrt(i)
}
toc()
```
